---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

ptstem
==========================================
> Stemming Algorithms for the Portuguese Language

[![Travis-CI Build Status](https://travis-ci.org/dfalbel/ptstem.svg?branch=master)](https://travis-ci.org/dfalbel/ptstem)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/dfalbel/ptstem?branch=master&svg=true)](https://ci.appveyor.com/project/dfalbel/ptstem)
[![Coverage Status](https://img.shields.io/codecov/c/github/dfalbel/ptstem/master.svg)](https://codecov.io/github/dfalbel/ptstem?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ptstem)](https://cran.r-project.org/package=ptstem)
[![](http://cranlogs.r-pkg.org/badges/ptstem)](http://cran.r-project.org/web/packages/ptstem/index.html)

This packages wraps 3 stemming algorithms for the portuguese language available in R. It
unifies the API for the stemmers and provides easy stemming completion.

## Installing

`ptstem` is not yet available in CRAN. You can install directly from Github using:

```{r, eval = F}
devtools::install_github("dfalbel/ptstem")
```

## Using

Consider the following text, extracted from [Stemming in Wikipedia](https://pt.wikipedia.org/wiki/Stemiza%C3%A7%C3%A3o)

```{r}
text <- "Em morfologia linguística e recuperação de informação a stemização (do inglês, stemming) é
o processo de reduzir palavras flexionadas (ou às vezes derivadas) ao seu tronco (stem), base ou
raiz, geralmente uma forma da palavra escrita. O tronco não precisa ser idêntico à raiz morfológica
da palavra; ele geralmente é suficiente que palavras relacionadas sejam mapeadas para o mesmo
tronco, mesmo se este tronco não for ele próprio uma raiz válida. O estudo de algoritmos para
stemização tem sido realizado em ciência da computação desde a década de 60. Vários motores de
buscas tratam palavras com o mesmo tronco como sinônimos como um tipo de expansão de consulta, em
um processo de combinação."
```

This will use the [`rslp`](https://github.com/dfalbel/rslp) algorithm to stem the text.

```{r}
library(ptstem)
ptstem(text, algorithm = "rslp", complete = FALSE)
```

You can complete stemmed words using the argument `complete = T`.

```{r, eval = F}
ptstem(text, algorithm = "rslp", complete = TRUE)
```

The other implemented algorithms are:

* hunspell: the same algorithm used in OpenOffice corrector. (available via [hunspell](https://github.com/ropensci/hunspell) package)
* porter: available via SnowballC package.

You can stem using those algorithms by changing the `algorithm` argument in `ptstem` function.

```{r}
library(ptstem)
ptstem(text, algorithm = "hunspell")
ptstem(text, algorithm = "porter")
```
