---
title: "ptstem: Stemming Algorithms for the Portuguese Language"
author: "Daniel Falbel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ptstem: Stemming Algorithms for the Portuguese Language}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> In linguistic morphology and information retrieval, stemming is the process of
> reducing inflected (or sometimes derived) words to their word stem, base or 
> root form—generally a written word form. The stem need not be identical to the 
> morphological root of the word; it is usually sufficient that related words 
> map to the same stem, even if this stem is not in itself a valid root. 
> Algorithms for stemming have been studied in computer science since the 1960s. 
> Many search engines treat words with the same stem as synonyms as a kind of 
> query expansion, a process called conflation.

From [Wikipedia](https://en.wikipedia.org/wiki/Stemming)

This paragraph gives a nice explanation of what *stemming* is. Much of academic 
work on stemming was focused on English Language and it's somewhat hard to find
stemming algorithms for other languages. `ptstem` tries to fix this, by 
providing a comprehensive interface for Portuguese Language stemming algorithms. 

The implemented algorithms are:

- **rslp**: [RSLP](http://doi.ieeecomputersociety.org/10.1109/SPIRE.2001.10024) is a simple and effective suffix-stripping algorithm for Portuguese. Implemented in `R` in the [`rslp`](https://CRAN.R-project.org/package=rslp) package.
- **hunspell**: [Hunspell](https://hunspell.github.io/) is the spell checker of LibreOffice, OpenOffice.org, Mozilla Firefox 3 & Thunderbird, Google Chrome, and it is also used by proprietary software packages, like Mac OS X, InDesign, memoQ, Opera and SDL Trados. It's implemented in `R` in the [`hunspell`](https://CRAN.R-project.org/package=hunspell) package.
- **porter**: The [Porter stemming algorithm](https://tartarus.org/martin/PorterStemmer/) (or ‘Porter stemmer’) is a process for removing the commoner morphological and inflexional endings from words. In `R` it's implemented in the [`SnowballC`](https://CRAN.R-project.org/package=SnowballC) package.


## Stemming

`ptstem` has only one important function that is called `ptstem`. You can easily 
stem a text by passing it to `ptstem`.

```{r}
library(ptstem)
```

### Example 1

```{r}
text <- "Em morfologia linguística e recuperação de informação a stemização (do inglês, stemming) é
o processo de reduzir palavras flexionadas (ou às vezes derivadas) ao seu tronco (stem), base ou
raiz, geralmente uma forma da palavra escrita. O tronco não precisa ser idêntico à raiz morfológica
da palavra; ele geralmente é suficiente que palavras relacionadas sejam mapeadas para o mesmo
tronco, mesmo se este tronco não for ele próprio uma raiz válida. O estudo de algoritmos para
stemização tem sido realizado em ciência da computação desde a década de 60. Vários motores de
buscas tratam palavras com o mesmo tronco como sinônimos como um tipo de expansão de consulta, em
um processo de combinação."
```

```{r}
ptstem(text)
```

By default `ptstem` uses the *rslp* algorithm to stem, and it complete stems with 
the most frequent word in the text (This is explained later). Is this example 
it's a little hard to see improvements with stemming, because the text doesn't 
contain many words with the same root. Let's look at a more simple example.

### Example 2

```{r}
text <- c("avião", "aviões", "aviação", "viação", "aves", "balão", "balões")
ptstem(text)
```

You can return the suffix stripped words (without completion) by setting the
argument `complete = FALSE`.

```{r}
ptstem(text, complete = FALSE)
```

You can also change the algorithm used to stem by setting the `algorithm` argument.

```{r}
ptstem(text, algorithm = "hunspell", complete = FALSE)
```

The hunspell stemmer is not a suffix-stripping algorithm, so it can find related
words that has the same sufffix. It happened here with the word "aviação" that 
was related to "viação" instead of "avião" and "aviões". Also you can see that 
hunspell is returning valid words, even with `complete = FALSE`, but it does not 
necessarily returns words that appear in the text, see:

```{r}
ptstem("aviões", "hunspell", complete = FALSE)
```

To use the Porter stemmer, simply tweak the `algorithm` argument again.

```{r}
ptstem(text, algorithm = "porter", complete = FALSE)
```

As Porter stemmer, is a general algorithm, it has some problems when detecting 
irregular forms of words. In this example, the stemming didn't relate any words, 
if you hadn't used the `complete = FALSE` argument, you wouldn't have noticed any
difference between the input and the output vectors.

```{r}
ptstem(text, algorithm = "porter")
```

### Example 3

`ptstem` has two other arguments that can be used to ignore words in stemming.

- `n_char`: minimum number of characters of words to be stemmed
- `ignore`: vector of words and regex's to igore

Sometimes you have some words in a text that you don't want to stem, like 
proper names or words in other languages and it's usefull to ignore them. 
Sometimes you also have very small words, that if stemmed they loose their 
meaning, the `rslp` algorithm has some rules about words lenghts, but `hunspell`
does not. That's why `n_char` argument is available.

```{r}
text <- c("obama", "gostei", "gostou", "gostamos", "é", "e")
```

```{r}
ptstem(text, complete = FALSE)
```

Here `rslp` stemmed "obama" to "obam" and "firmware" to "firmw". You can choose
to not stem theese words by setting the `ignore` parameter.

```{r}
ptstem(text, complete = FALSE, ignore = c("obama"))
```

By default, `ptstem` does not stem words with less then three characters. If you
set for at least 1 characters.

```{r}
ptstem(text, complete = FALSE, n_char = 1)
```

You can see that "e" and "é" were united.
It's also possible to ignore regex's, using the `ignore` argument.

```{r}
ptstem(text, complete = FALSE, ignore = c("go."))
```

This doesn't stem words that start with "go".

## Performance

The goal of stemming algorithms is to group related words and to separate unrelated words. With this in mind, you can talk about two kinds of possible errors when stemming:

* Understemming: Related words were not grouped because you didn't stem enought.
* Overstemming: Unrelated words were grouped because you removed a large part of the word when stemming.

To measure these errors the function `performance` was implemented. It returns a `data.frame` with 3 columns. The name of the stemmer and 2 metrics:

* UI: the undersampling index. It's the proportion of related words that were not grouped.
* OI: the overstemming index. It's the proportion of unrelated words that were grouped. 

Remember that OI is 0 if you don't stem. So I think the true objective of a stemming algorithm is to reduce UI without augmenting OI too much.

`ptstem` package provides a dataset of grouped words for the portuguese language (found in this [link](http://www.inf.ufrgs.br/~fnflores/paice_tool/)). It's in this dataset that `performance` function calculates the metrics described above.

See results:

```{r}
performance()
```

This is not the only approach for measuring performance of the those algorithms. The article [*Assessing the impact of Stemming Accuracy on Information Retrieval – A multilingual perspective*](http://dx.doi.org/10.1016/j.ipm.2016.03.004) describes various ways to analyse stemming performance.
